\documentclass{report}

\usepackage{listings}
\usepackage[vlined, boxed]{algorithm2e}
\usepackage{tikz}
\usepackage{hyperref}
\usepackage{float}
%\usepackage{geometry}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{galois}
\usepackage{titlesec}
\usepackage{stmaryrd}
\usepackage{tabularx}
\usepackage{fancyvrb}
\usepackage{enumitem}

\titleformat{\chapter}[hang]
  {\normalfont\huge\bfseries}   % formatting for chapter title (font size, style)
  {\thechapter.}                % the label (chapter number followed by a period)
  {1em}                         
  {}
\titlespacing*{\chapter}{0pt}{0pt}{20pt}

\SetAlCapSkip{2mm}

\newcommand\nElement{\mathrlap{\perp}\top}

\title{Implementation and Evaluation of the FunArray\\[1em]\large{}Bachelor's Thesis\\[1em]Software and Computational Systems Lab\\Ludwig-Maximilians-Universit\"at M\"unchen}
\date{October 2024}
\author{Maximilian Hofstetter}
%\institute{Software and Computational Systems Lab\\Ludwig-Maximilians-Universit\"at M\"unchen}
%\subtitle{Bachelor's Thesis}

\newcommand{\funArray}[1]{$#1$}
\newcommand{\bound}[1]{\{#1\}}
\newcommand{\fvalue}[1]{\;#1\;}



\begin{document}

\section*{Statement of Originality}

I hereby confirm that I have written the accompanying thesis by myself, without contributions from any sources other than those cited in the text and acknowledgements. This applies also to all graphics, drawings, maps and images included in the thesis.

\section*{Utilised Aids in Creating this Thesis}

When writing source code for my thesis, I utilised the IntelliJ IDEA full line completion\footnote{\url{https://www.jetbrains.com/help/idea/full-line-code-completion.html}}. This tool suggests possible code fragments to the developer based on a deep learning model. At the time of the creation of this thesis it is not yet able to suggest any in-depth design choices. I did not use any other AI tools for writing source code nor text in this paper.

\maketitle
\setcounter{tocdepth}{1}
\tableofcontents

\chapter{Introduction}
\chapter{Background}
\section{Program semantics}

Consider the following computer program written in some simple fictive programming language.

\begin{center}
\begin{BVerbatim}
i := 0;
while (i < 10) {
    i := i + 1;
}
\end{BVerbatim}
\end{center}

An experienced programmer might be quick to think they know what this code is supposed to do. They however can only see the syntax of this program and have no knowledge of what the statements actually do in the made-up language. To discover what is happening we first must define the language's semantics. 


\subsection{Environments and Expressions}

Determining the meaning of a particular statement requires us to know the current state of the program. The expression \texttt{i + 1} has an entirely different result, depending on the value of the variable \texttt{i}, which is recorded in the program's concrete scalar variable environment. An environment is defined as a function $\rho\in\mathcal{R}_v$ with $\mathcal{R}_v \triangleq \mathbb{X}\mapsto\mathcal{V}$, where $\mathbb{X}$ is the set of syntactical variable names and $\mathcal{V}$ is the set of values they can take \cite{cousot2011}.

Now consider the simple syntactical expression $\mathtt{e}\in\mathbb{E}$. It only contains scalar constants, scalar variable references and operators. Now $\llbracket\mathtt{e}\rrbracket\rho$ describes the semantics of \texttt{e} in the concrete variable environment $\rho$. The semantics of a variable \texttt{i} is defined as $\llbracket\mathtt{i}\rrbracket\rho=\rho(\mathtt{i})$ and the semantics of a syntactic operator $\circledast$ is defined as $\llbracket\mathtt{e_0\circledast e_1}\rrbracket\rho=\llbracket\mathtt{e_0}\rrbracket\rho \ast\llbracket\mathtt{e_1}\rrbracket\rho$ \cite{scott1971}. 
Imagine a variable environment $\rho$, where $\rho(\mathtt{i})=1$, for example. Our previously mentioned expression \texttt{i + 1} evaluates to:

\begin{equation*}
\begin{aligned}
\llbracket\mathtt{i \;\texttt{+}\; 1}\rrbracket\rho &=\llbracket\mathtt{i}\rrbracket\rho +\llbracket\mathtt{1}\rrbracket\rho \\
& = \rho(\mathtt{i}) + 1\\
& = 1+1\\
& = 2
\end{aligned}
\end{equation*}


\subsection{Transformers}\label{transformers}
A computer program not only contains expressions, but also commands or statements that modify the environment itself. For example, consider the environment $\rho$ where $\rho(\mathtt{i})=1$. After executing the command \texttt{i := i + 1}, the value of the variable \texttt{i} changes and therefore the environment $\rho'$ after the execution of the command is different from $\rho$. We write $\llbracket \texttt{i\;:=\;i\;+\;1} \rrbracket\rho=\rho[\mathtt{i}:=\texttt{i\;+\;1}]=\rho'$, meaning the semantics of the environment $\rho$ after executing an assignment \texttt{i\;:=\;i\;+\;1} is an environment $\rho'$ where all occurrences of \texttt{i} have been replaced by \texttt{i\;+\;1}.

More generally the semantics of an assignment is defined as $\llbracket \texttt{i\;:=\;e} \rrbracket\rho=\rho[\mathtt{i}:=\texttt{e}]$ with $\rho[\mathtt{i}:=\texttt{e}](\mathtt{i})=\llbracket\mathtt{e}\rrbracket\rho$ \cite{cousot2011}. Commands that alter the environment are called transformations. They are basically functions in $\mathcal{R}_v\mapsto\mathcal{R}_v$ \cite{scott1971}.


\section{Abstract Interpretation}

Let us take a look at the equation $8174+2938=x$. We want to determine whether $x$ has a positive or a negative value. Instead of doing mental arithmetic and actually adding $2938$ to $8174$ we almost instinctively know that $x$ must be positive because both its summands are positive. Without realising it, we abstracted the operands of the equation to its sign and calculated it with only this property. When applied to static code analysis, this strategy we just instinctively applied, is called abstract interpretation. Its main idea is that for proving a program's properties it is not necessary to run it with concrete values. Instead it can be run on the examined properties \cite{cousot2023, cousot1977}.


\subsection{Properties}

Consider a set of entities $\mathcal{E}$. A formal property $P\subseteq \wp(\mathcal{E}) $ is a set of entities that have this property \cite[chapter 8]{cousot2021}. With the set of integers as our set of entities $\mathcal{E}=\mathbb{Z}$, our ``positive'' property is defined as $P_{pos}=\{z\in \mathcal{E} \;|\; 0<z\}$ and we can define some more semantic properties like in table \ref{table:properties}.
\begin{table}[hbt]
\begin{center}
  \begin{tabular}{l|l}
  $\mathsf{false}$ & $\emptyset$\\
   $P_{0}$ & $\{0\}$\\
   $P_{pos}$ & $\{z\in \mathbb{Z} \;|\; 0<z\}$\\
   $P_{neg}$ & $\{z\in \mathbb{Z} \;|\; 0>z\}$\\
   $P_{pos0}$ & $\{z\in \mathbb{Z} \;|\; 0\leq z\}$\\
   $P_{neg0}$ & $\{z\in \mathbb{Z} \;|\; 0\geq z\}$\\
   $P_{even}$ & $\{z\in \mathbb{Z} \;|\; \exists k\in\mathbb{Z}:2k=z \}$ \\
   $P_{odd}$ & $\{z\in \mathbb{Z} \;|\; \exists k\in\mathbb{Z}:2k+1=z \}$ \\
   $P_{pos\_even}$ & $\{z\in \mathbb{Z} \;|\; z>0 \wedge \exists k\in\mathbb{Z}:2k=z \} = P_{pos} \cap P_{even}$ \\
   $\vdots$ &$\vdots$\\
   $\mathsf{true}$ & $\mathbb{Z}$\\
  \end{tabular}
  \caption{An incomplete selection of properties for $\mathbb{Z}$.}\label{table:properties}
  \end{center}
\end{table}

We call a property $P$ ``stronger'' than $P'$ (or $P'$ ``weaker'' than $P$) if $P\subseteq P'$. Its set representation contains less elements and we therefore have more information about an element that fulfils said property. The weakest property is the $\mathsf{true}$ property that is satisfied by alle elements and the strongest is $\mathsf{false}$ which is satisfied by none \cite[chapter 8]{cousot2021}. Not all properties from $\wp(\mathcal{E})$ carry a semantic meaning. We therefore only consider the set of properties of interest $C\subseteq\wp(\mathcal{E})$. If we want to analyse the signs of integers this could be $C_{sign}=\{\mathsf{false}, P_{0}, P_{pos}, P_{neg}, P_{pos0}, P_{neg0}, P_{not0}, \mathsf{true}\}$. The poset  $\langle C,\subseteq\rangle$ is called the ``concrete domain''.

\subsection{Abstraction}\label{sec:abstractinterpretation:abstraction}

Now consider a set of abstract properties $A$. This is not a power set of concrete values anymore but a set of purely semantic symbols that represent abstract properties. If there exists a Galois connection $\langle C,\subseteq\rangle \galois{\alpha}{\gamma}\langle A\sqsubseteq\rangle$, we call $\langle A\sqsubseteq\rangle$ the abstract domain, $\alpha\in C\to A$ the abstraction function and $\gamma\in A\to C$ the concretisation function \cite[chapter 11]{cousot2021}. To put the working principle of Galois connections into simple terms: $A$'s elements abstract $C$'s elements and $\sqsubseteq$ works analogous in the abstract to $\subseteq$ in the concrete. More formally: $\forall P_C\in C:\forall P_A \in A:\alpha(P_C)\sqsubseteq P_A \Longleftrightarrow P_C \subseteq \gamma(P_A)$.

\input{res/hasse_sign.tex}

\subsection{Property Transformers}

When executing a computer program in the concrete, there are transformer semantics in 

\begin{table}[hbt]
\begin{center}
\begin{tabular}{c|c|c|c|c|c|c|c|c}
            $\times_\pm$& $\perp$ & $<0$    & $0$     & $>0$    & $\leq0$ & $\neq0$ & $\geq0$ & $\top$  \\ \hline
            $\perp$ & $\perp$ & $\perp$ & $\perp$ & $\perp$ & $\perp$ & $\perp$ & $\perp$ & $\perp$ \\
            $<0$    & $\perp$ & $>0$    & $0$     & $<0$    & $\geq0$ & $\neq0$ & $\leq0$ & $\top$  \\
            $0$     & $\perp$ & $0$     & $0$     & $0$     & $0$     & $0$     & $0$     & $0$     \\
            $>0$    & $\perp$ & $<0$    & $0$     & $>0$    & $\leq0$ & $\neq0$ & $\geq0$ & $\top$  \\
            $\leq0$ & $\perp$ & $\geq0$ & $0$     & $\leq0$ & $\geq0$ & $\top$  & $\leq0$ & $\top$  \\
            $\neq0$ & $\perp$ & $\neq0$ & $0$     & $\neq0$ & $\top$  & $\neq0$ & $\top$  & $\top$  \\
            $\geq0$ & $\perp$ & $\leq0$ & $0$     & $\geq0$ & $\leq0$ & $\top$  & $\geq0$ & $\top$  \\
            $\top$  & $\perp$ & $\top$  & $0$     & $\top$  & $\top$  & $\top$  & $\top$  & $\top$ 
        \end{tabular}
  \caption{A table showcasing the result of the abstract property transformer $\times_\pm$ which is an abstraction of the concrete multiplication operator.}\label{table:multiply}
  \end{center}
\end{table}

\subsection{Fixpoint analysis}

\subsection{The Interval Domain}
asdf

\input{res/hasse_interval.tex}

\section{FunArray}
\subsection{Functors}
\subsection{Abstractions}
\subsection{Segmentation Functor: The FunArray}




\chapter{Implementation}



\section{Scalar Variables}\label{sec:scalarvariables}
Cousot, Cousot and Logozzo's FunArray is embedded into Clousot, an already existing static code contract analyser for the .NET framework \cite{cousot2011}. It therefore profits from an already existing abstract analysis for scalar variables. My implementation however works in a rudimentary standalone analyser and implements its own abstract domains for scalar variables.
They are defined by the interface \texttt{DomainValue} and support only a limited amount of arithmatic operations: Addition, subtraction, multiplication, division, modulo, negation and absolute value. Operations like exponentiation or bitwise xor etc. that are available in many programming languages like C or Java, have not been implemented, which limits the amount of programs that can be analysed with my implementation. 

Since the focus of my work is to test the FunArray technique I have refrained from implementing more complicated abstract domains (that would have delivered more precise results), like for example the relational approach of polyhedra \cite{cousot1978}. Instead I primarily focused on the interval domain. To prove that the FunArray works with different scalar variable abstractions I also partially implemented the sign domain.

\subsection{Intervals}

I chose to only abstract integers with this domain, as the use of floating point numbers and open intervals would introduce unnecessary inaccuracies and the benchmark I later tested against only included integers as well. The implementation of the interval domain is relatively straight-forward. An interval $[a,b]$ consists of a lower limit $a$ and an upper limit $b$ with $a,b \in \mathbb{Z}\cup\{-\infty,\infty\}$. The set of $\mathbb{Z}\cup\{-\infty,\infty\}$ is implemented by the \texttt{InfInt} class. It allows calculations on infinities, such as $\infty - 10 = \infty$ or $(-4) \cdot \infty = -\infty$, by extending the operations of arithmetics in $\mathbb{Z}$. It is assumed that there are no integer overflows.


\subsubsection{Arithmetic Expression Operations}

Table \ref{table:intervalarithmetics} shows the supported expression arithmetics for the \texttt{ReachableInterval} class and how they are implemented. Most operations are intuitive and follow the classical interval arithmetic \cite{dawood2011}. Operations with at least one of the operands being of the class \texttt{Unreachable} representing $\bot$, yield another \texttt{Unreachable}. 

The abstraction of the modulo operation is not as straight-forward as the others. Different programming languages define their modulo operator differently. For example \texttt{7 \% -5} in C or Java yields a result of $2$ whereas the same expression in Python yields $-3$. Both results are mathematically correct, because $2 \equiv -3 \mod -5$, when talking about equivalence classes. The programming languages just translate those equivalence classes differently into integers. I chose for my interval domain to follow the semantics of C, where the sign of the result of a modulus operation is equal to the dividend's sign. The modulo operation also over-approximates the result for ease of implementation.

\input{res/interval_arithmetics}



\subsubsection{Abstract operations}

Table \ref{table:intervalabstractoperations} shows the implementations of the abstract operations for the \texttt{Interval} class. The join $\sqcup$ and meet $\sqcap$ are basically equivalent to their concrete set operations $\cup$ and $\cap$. When the result of a meet operation $[\underline{x},\overline{x}] \sqcap [\underline{y},\overline{y}] = [\underline{z},\overline{z}]$ would be an illegal interval, such that 
$\underline{z}>\overline{z}$, instead an \texttt{Unreachable} is returned.

\input{res/interval_abstract_operations}

\subsubsection{Abstraction and Concretisation Functions}

The abstraction function defined in the \texttt{Domain} deviates from the mathematical definition as explained in section \ref{sec:abstractinterpretation:abstraction}. Instead of abstracting from a property (a set of integers), it abstracts from a single integer, since a concrete program only deals with integers but not with sets. The abstraction function is needed when evaluating an expression containing constants as described in section \ref{sec:expressions:evaluation}. It is implemented as $\alpha(z)=[z,z]$.
The concretisation function $\gamma$ deviates in the same way and can only concretise intervals $\gamma([\underline{z},\overline{z}])=z$ where $z = \underline{z} = \overline{z}$. When $\underline{z} \neq \overline{z}$, the method throws a \texttt{ConcretisationException}. The concretisation function is needed for normalising expressions as described in section \ref{sec:expressions:normalisation} and the exception is being handled in that process. 

\subsection{Signs}
I also partially implemented the Sign domain to make sure that my implementation of the FunArray works as intended with different abstractions of element values. The \texttt{Sign} class however does not implement all methods defined in the \texttt{DomainValue} interface and cannot be used in a full analysis of a program.
\section{Arrays}

The \texttt{FunArray} class provides a collection of bounds, element values and segment emptiness information. I decided against implementing the FunArray as a linked list and instead stored these values in Java collections as to make use of the Java stream API. The emptiness information is stored as boolean values (a segment can either be might-be-empty or non-empty). Array elements can take any implementation of the \texttt{DomainValue} (See section \ref{sec:scalarvariables}). Bounds are stored as \texttt{Bound} objects. The \texttt{FunArray} is immutable and when changed provides modified copies of itself.

Apart from some minor utility methods, the \texttt{FunArray} provides four major methods: \texttt{get}, \texttt{insert}, \texttt{join} and \texttt{widen}. The \texttt{get} method takes a normal expression as an argument and returns the element value at that index. Since that very index might not necessarily be present in the bounds, this operation is not as trivial as it sounds. The array is traversed from the left and the rightmost bound that can be said to be less than the given index is being searched. After that, the array is traversed from the right and the leftmost bound that is greater than the index plus one is found. The result of the \texttt{get} operation then is the join of all array elements that lie between both these bounds. If it is not possible to find bounds less or greater than the given index, instead the most outer bounds of the array are being used, as it is simply assumed that no array access is ever out of bounds. The \texttt{insert} method works in a similar way. Since the desired index might not actually be present in the bounds of the array, it is not possible to directly insert the designated value. Instead, once again, the rightmost bound smaller than the index and the leftmost bound greater than the index are being searched. The bounds between them are being removed and the values between them are joined. Then a new segment with the desired value is placed in the middle of this joined segment \cite{cousot2011} . Consider an Array $A$, looking like $B_1P_1B_2[?_1]\ldots P_{n-1}B_n[?_{n-1}]$, where $B_n$ is the $n$-th bound, $P_n$ is the $n$-th value and $[?_{n}]$ is the emptiness flag for the $n$-th segment. We want to insert a value $v$ at the index $i$. The rightmost bound smaller than the index $B_l$ and the leftmost bound greater than the index $B_g$ are determined and the segments in between are being joined: $B_1P_1B_2[?_1]\ldots B_l[?_l] (\bigsqcup^{g-1}_{k=l}P_k) B_g[?_g]\ldots P_{n-1}B_n[?_{n-1}]$. Then the desired value  is inserted in this joined segment like this: $B_1P_1B_2[?_1]\ldots\allowbreak B_l[?_l] (\bigsqcup^{g-1}_{k=l}P_k)\allowbreak\, \{\,i\,\}?\, v \,\{\,i+1\,\}\, \allowbreak(\bigsqcup^{g-1}_{k=l}P_k) B_g?\ldots \allowbreak P_{n-1}B_n[?_{n-1}]$. Edge cases in which for example $i\in B_l$ are being handled, but won't be described any further here. The \texttt{join} and \texttt{widen} operations first execute a unification as described in section \ref{sec:funarray:unifying} and then join or widen the element values segment wise.

\textcolor{red}{Evtl diesen Part eher in den Background packen. Ist zwar so auch implementiert, aber ist eigentlich nicht von mir ausgedacht worden. Hier in dieser Section nur drauf eingehen, wies denn eigentlich implementiert wurde. Bzw evtl diesen Part weiter nach hinten verschieben in die Assignment Section}

\subsection{Bounds and Normal Expressions}

Expressions that appear in the bounds of a FunArray follow a chosen normal form. These are implemented separately from the arithmetic program expressions described in section \ref{sec:expressions} in the class \texttt{NormalExpression}. This split design was chosen because the bound expression need to fulfil some additional functionality like syntactic comparison. I adapted Cousot, Cousot and Logozzo's normal form of $\mathtt{v}+k$, where \texttt{v} is a \texttt{Reference} and $k$ is a constant integer value \cite{cousot2011}. The \texttt{Reference} interface is implemented by the \texttt{VariableReference} class, which is nothing more than a wrapper for a string, representing a variable name, and the \texttt{ZeroReference}, which is a special variable whose value is assumed to always be zero, such that it is possible to represent constants as normal expressions. Two normal expressions can be syntactically compared if their variable reference is the same, by simply comparing their constant value. A refinement that has not been made in my implementation would be to semantically compare bound expressions. This syntactical comparison is needed when getting or inserting a value into a FunArray.

The \texttt{Bound} class is a wrapper for a set of \texttt{NormalExpression} that provides some set algebra (union, intersection etc.) as this is not part of the standard Java set. It also provides functionality to modify the bound accordingly when a variable value changes (a new expression needs to be added, when a variable value is set to a value already present in that bound for example).

\subsection{Unifying two FunArrays}\label{sec:funarray:unifying}

Cousot, Cousot and Logozzo developed a recursive algorithm for unifying two FunArray segmentations \cite{cousot2011}. It is subjectively complicated to understand at first glance and needs to handle 11 different distinct sub-cases. It is a recursive algorithm that relies on the assumption that the left side of the array is already unified, but in one case that assumption is violated, making it hard to actually implement it recursively. When I implemented it, it was also prone to errors and hard to debug. Therefore I chose to implement the unifying of two segmentations in a different iterative algorithm. The authors also mentioned that they refined their algorithm in their actual implementation of the FunArray, mentioning a ``purification step'' that sounds similar to my approach, but don't go into more detail \cite[section 12.2]{cousot2011}. As their implementation is not freely available, one could only guess that they might have actually taken a similar approach as me in the end. My algorithm works as follows for two FunArrays $A$ and $B$ and their respective neutral elements $\nElement_A$ and $\nElement_B$:
\begin{enumerate}
  \item  Remove all expressions in the bounds of $A$ and $B$ that are only present in either of them but not both.
  \item \label{item:removeemptybounds} If that leaves empty bounds, such that there is no expression in them left, remove that bound and merge the adjacent segment values via the join operation. Set the might-be-empty flag for that new segment only if both its predecessors had it.
  \item Now traverse the bounds of both FunArrays from left to right. Treat the bounds as sets of expressions. The current bounds are denoted as $B_A$ and $B_B$.
  \begin{enumerate}[label*=\arabic*.]
    \item If $B_A \cup B_B = \emptyset$, replace $B_A$ and $B_B$ with empty bounds and go on to the next bounds.
    \item If $B_A \cup B_B \neq \emptyset$, add a bound $B_A \setminus B_B$ to the right of $B_A$. Add an element value $\nElement_A$ in between them and set the might-be-empty flag to true. Add a bound $B_B \setminus B_A$ to the right of $B_B$ analogously. Replace both $B_A$ and $B_B$ with $B_A \cup B_B$. Go on to the newly created bounds to the right.
  \end{enumerate}
  \item Finally remove all empty bounds in the resulting FunArrays, as described in Step~\ref{item:removeemptybounds}.
\end{enumerate}

\noindent\textcolor{red}{Frage @Gidon: sollte ich hier irgendwie beweisen, dass der Algorithmus ein korrektes Ergebnis liefert?}
\section{States}
\section{Arithmetic Expressions}\label{sec:expressions}

Arithmetic expressions are defined in the \texttt{Expression} interface. I implemented the three atomic expressions \texttt{Constant}, \texttt{Variable} and \texttt{ArrayElement} and the five expression operations \texttt{Addition}, \texttt{Subtraction}, \texttt{Multiplication}, \texttt{Division} and \texttt{Modulo}. 

\subsection{Evaluation of an Expression}\label{sec:expressions:evaluation}
Evaluating an expression is basically the same as determining its abstract semantics under the current state. Since the \texttt{DomainValue} class has the five arithmetic operations already defined as methods, the evaluation implementation of the expression operations is obvious. The evaluation of the three atomic expressions however is not as simple. Since the FunArray allows differing value domains for variables and its elements, it might not possible to calculate with those straight-away. The constant expression represents a concrete constant value and therefore lies in yet another domain. Cousot, Cousot and Logozzo chose to evaluate their expressions in the variable domain \cite{cousot2011}. There is no immediately obvious reason as to why an evaluation in the array element domain should not be considered, but I chose to follow their lead nevertheless. Therefore, two translation functions are needed to translate the concrete constant values and the array element values into the appropriate value in the variable value domain. The former is simply the abstraction function $\alpha$ which is provided by the \texttt{Domain} class and the latter called \texttt{convert\allowbreak{}Array\allowbreak{}Element\allowbreak{}Value\allowbreak{}To\allowbreak{}Variable\allowbreak{}Value} is provided by the \texttt{AnalysisContext}interface. When both variables and array elements are being abstracted into the same domain, this is simply the identity function. Its opposite, \texttt{convert\allowbreak{}Variable\allowbreak{}Value\allowbreak{}ToArray\allowbreak{}Element\allowbreak{}Value}, is needed when the calculated value of an expression is supposed to be assigned to an array element.


\subsection{Normalisation of an Expression}\label{sec:expressions:normalisation}

Since general expressions (that are not already in normal form) are being used as indeces when assigning a value to an array element, they need to be able to be brought into normal form. This works differently for different expressions. Some expressions may be able to be brought into more than one normal form or none at all. Consider the expression \texttt{e = a + b} and the state $\rho=\{\mathtt{a}\mapsto[3,3], \mathtt{b}\mapsto[7,7]\}$ for example. There are actually 3 ways \texttt{e} can be written in our $\mathtt{v}+k$ normal form: $\mathtt{a}+7$, $\mathtt{b}+3$ or $\mathtt{v_0}+10$, where $\mathtt{v_0}$ is a special variable whose value is always assumed to be zero. All three of these are being considered when inserting a value into an array.

\subsubsection{Addition and Subtraction}


Consider an addition expression $\mathtt{a}=\mathtt{e}_1 + \mathtt{e}_2+\ldots+\mathtt{e}_n$. Its normal forms are derived the following way:


\begin{enumerate}
	\item Consider a single normal form $\mathtt{v}_i + k_i$ of a single summand $\mathtt{e}_i$.
	\item Evaluate the expression $\mathtt{a}'=\mathtt{e}_1 +\ldots+ \mathtt{e}_{i-1} + \mathtt{e}_{i+1} +\ldots+\mathtt{e}_n+k_i$.
	\item If its value is concretisable, meaning it can be concretised into a single integer value (e.g. $\gamma([3,3])=3$), add $\mathtt{v}_i+\gamma(\llbracket\mathtt{a}'\rrbracket)$ to the list of normal forms for \texttt{a}.
	\item Repeat for all operands \texttt{e} in \texttt{a}. 
\end{enumerate}

\noindent The normalisation of subtraction works the same way, since a subtraction is nothing more than an addition with exactly two operands where the value of the second operand is being inverted when evaluating.

\subsubsection{Variable and Constant}
Normalising a constant expression \texttt{k} is trivial, as its normal form is simply $\mathtt{v_0}+k$, where $\mathtt{v_0}$ is a special variable whose value is assumed to always be zero of the class \texttt{ZeroReference}. Equally as trivial is the normalisation of a variable expression \texttt{v}. Its normalisation is $\mathtt{v}+0$. If $\llbracket\mathtt{v}\rrbracket$ is concretisable, meaning it describes a single concrete value $k'$, the second normal form of \texttt{v} is $\mathtt{v_0}+k'$.


\subsubsection{Multiplication, Division, Modulo and Array Element}

The normalisation of the multiplication, division and modulo expressions is quite similar to the constant normalisation. As there es no general way to rearrange them into normal form, they can only be normalised, when their evaluation can be concretised into the constant part of the normal form. Of course there are certain refinements that could be done here, like rearranging multiplications into additions when a factor can be concretised. These opportunities have not been seized to keep the process of normalisation as simple as possible and as these refinements would probably only yield a marginal reward. 

\section{Assignments}

\section{Conditionals}\label{sec:conditions}

When defining the semantics of a real world programming language like C, we would normally not only need to assign semantics to arithmetic expressions like in the previous section, but also to boolean expressions \cite{scott1971}. As my analyser is not going to be used to analyse C programs or programs in any other real world programming language, I skipped this step. 




\section{Loops}


\chapter{Evaluation}
\textcolor{red}{-- Forschungsfragen nochmal aufgreifen}

To evaluate 

\section{Experimental Setup}
\subsection{Benchmarks}

The International Competition on Software Verification (or SV-COMP)\footnote{More information about SV-COMP available at \url{https://sv-comp.sosy-lab.org/}} provides a set of benchmarks that can be used to compare tools for software verification\footnote{Available at \url{https://gitlab.com/sosy-lab/benchmarking/sv-benchmarks}}. Their subcategory \textit{c/ReachSafety-Arrays} contains a collection of 433 verification tasks written in C for which the treatment of arrays is necessary to determine reachability. As such, this collection is suitable to test my implementation of the FunArray. They include at least one assertion per benchmark and its expected verdict for testing 
The simple C programs have been translated to the embedded domain-specific language in Java using the tool Cuv\'ee \cite{ernst2020}. Since my implementation of the FunArray does not provide the entire functionality of the C language, about half of the benchmarks could not be translated. The 196 untranslatable benchmarks utilise features like the use of boolean variables, nested arrays, functions with a return value, structs and the use of the break keyword.

\subsection{Formal Arguments of the FunArray}

One of the FunArray's strengths is its ability to use different abstract domains for its bounds, array elements and variables without having to adapt the static analyser. For this evaluation the following abstract domains have been chosen as the FunArray functor's formal arguments: For abstracting variables and array elements the interval domain has been chosen, since it is relatively simple to implement. The chosen expression domain and therefore the abstract representation of segment bounds follows the scheme used by the FunArray's inventors \cite[section 7.2]{cousot2011}. Expressions are limited to the normal form $v+k$ where $v$ is a variable (whose value is abstracted by intervals) and $k$ is an integer constant.

\section{Results}

The execution time of the benchmarks is almost negligible. To execute all 237 benchmarks it takes 253 milliseconds (Averaged value over 100 runs. The workbench used is a M2 Max MacBook Pro running OpenJDK 22). This comes down to about a millisecond per benchmark. This result was expected, since the FunArray does not handle disjunctions separately but through the use of empty segments. No combinatorial explosion can occur.

\subsection{Invariants}
Spread across the 238 benchmarks there are 947 loops. For 580 or 61.3\% of those my implementation of the FunArray was able to determine a non-trivial invariant. A non-trivial invariant is defined as a state in which at least one of the arrays contains a segment whose value is not $\top$.
This definition of non-triviality, however, does not take the FunArrays ability to depict multiple segments into account. If we were to treat all arrays as a single segment with a homogenous value across all elements, we could have gotten the same result and the same number of non-trivial invariants. We therefore narrow down the non-triviality criterion to those states containing at least one array with more than one segment and at least one segment with a value that is not $\top$. Now we arrive at 427 non-trivial invariants, which is still 45.0\%. 
If we were to increase the required number of segments to more than two, this number would drop down to 3, suggesting the FunArrays ability to depict a lot of segments cannot be utilised in actual code analyses. No calculated invariants had more than three segments.

\subsection{Verifying assertions}

My implementation is able to correctly identify the result of 24 out of the 247 assertions spread among 237 benchmarks or 9.7\% of assertions. When manually inspecting the program mechanism preceding these assertions in the benchmark, a single pattern can be identified that precedes the majority of them: Initialisation. The benchmark first initialises an array with a consistent value as its every element and immediately afterwards asserts that every element of the array has that specific value. Of the 24 correctly assessed assertions, 20 follow this pattern. The remaining four assertions follow a more complex program scheme, that involves conditions and loops.



\section{Discussion}
\subsection{Comparison with Cousot, Cousot and Logozzo's Results}

Cousot, Cousot and Logozzo evaluate their implementation of the FunArray by analysing the main libraries of the .NET framework and the implementation itself \cite{cousot2011}. The .NET libraries have been chosen for this since they contain a large amount of code and a vast selection of functionalities and techniques. They, however, lack an important aspect: The specification of expected properties. The authors are able to infer thousands of invariants but cannot determine whether they actually provide any useful information for the further execution of the code. To check against pre-specified properties they annotated their implementation of the FunArray with pre-conditions and proof obligations and ran it on itself. They were able to correctly assert 61 out of 1800 or 3.4\% of proof obligations when analysing the FunArray implementation using itself \cite{cousot2011}. This roughly matches my results of 9.7\%, considering the easy initialisation assertions make up a way greater amount of my test cases than would be present in a real world program.

\subsection{Limitations}

\subsubsection{Adjacent potentially empty segments}
When expanding a segment by inserting a value adjacent to it, formerly adjacent and potentially empty segments will be joined with their nearest neighbour until a non-empty segment is reached. Let us consider the FunArray \funArray{A:\bound{a} \fvalue{x} \bound{b} \fvalue{y} \bound{c}? \fvalue{z} \bound{d}} for example. It is unknown whether its second segment containing the value $y$ is empty or not and thus whether $b=c$ or $b<c$ is true.

When trying to expand the first segment by inserting its value $x$ at index $b$, the resulting FunArray could either be \funArray{\bound{a} \fvalue{x} \bound{b\;c} \fvalue{x} \bound{b+1}? \fvalue{z} \bound{d}}, if the $y$-segment was empty, or alternatively \funArray{\bound{a} \fvalue{x} \bound{b} \fvalue{x} \bound{b+1} \fvalue{y} \bound{c}? \fvalue{z} \bound{d}}, if it was non-empty. Since the position of the \funArray{\bound{c}}-bound cannot by decided, it must be discarded altogether. After the insertion, $A$ becomes \funArray{A':\bound{a} \fvalue{x} \bound{b} \fvalue{x} \bound{b+1}? \fvalue{y\sqcup z} \bound{d}} and all information concerning the variable $c$ is lost.

 Analyses of programs utilising segments in their arrays with adjacent and potentially empty segments have to discard non-trivial information. Thus the FunArray's approach for handling potentially empty adjacent segments limits the number of programs for which an analysis can achieve a non-trivial result.
 
 Consider Algorithm \ref{algo:dijkstra}, Dijkstra's Dutch national flag algorithm, for example. Even though the analysis correctly determines the Array to be described as $\bound{0} \fvalue{[-\infty, -1]} \bound{r}? \fvalue{\top} \bound{w}? \fvalue{[0, 0]} \bound{b}? \fvalue{[1, \infty]}  \bound{|A|}?$ after a single loop pass, it loses all its information in the second pass, because the determined segments are all potentially empty and extending their neighbours discards them altogether. After the second loop pass the analysis yields $\bound{0} \fvalue{\top} \bound{|A|}$. Basically no information can be gained   
 
\input{res/dijkstra_algorithm.tex}


\chapter{Conclusion}

Strength: Initialization of Arrays




\newpage
\bibliographystyle{apalike2}
\bibliography{Bibliography}

\end{document}